---
title: Maxent model of Northern pintail (Anus acuta)
author: Pin-Yuan Chen | Winter 2023
date: 1/24/2023
output: html_document
editor_options: 
  chunk_output_type: console
---

# Load packages and data
```{r}
#library(rgbif)
#library(maptools)
# install.packages('ENMeval')
#library(ENMeval)
#library(raster)
#library(dismo)
# install.packages('rJava')

#Sys.setenv('JAVA_HOME'="C:/Program Files/Java/jdk-19/")
#library(rJava)
#library(rethinking)
#library(predicts)

library(terra)
```

## GBIF data
```{r}
#sp <- geodata::sp_occurrence("Anas", "acuta", download=FALSE)
aacuta <- read.csv('aacuta_clean.csv')
head(aacuta)

# Check the spatial resolution of the data
aacuta$decimalLongitude[1:6]
aacuta$decimalLatitude[1:6]
```
Database for occurrence records of the selected species from the gbif database. The search for occurrences was limited to 20,000 records in California. Duplicate records and data without coordinates has been removed. The remained data points are 2848 obs. with  17 variables.

## Environmental data
```{r}

envs <- geodata::worldclim_global(var = "bio", res = 5, path=".")
b <- c('bio_1', 'bio_4', 'bio_5', 'bio_6', 'bio_12', 'bio_16', 'bio_17')
nms <- paste0("wc2.1_5m_", b)
envs <- envs[[nms]]
```

[WorldClim dataset (Hijmans 2005)](https://www.worldclim.org/data/bioclim.html)

* BIO1 = Annual Mean Temperature
* BIO4 = Temperature Seasonality (standard deviation ×100)
* BIO5 = Max Temperature of Warmest Month
* BIO6 = Min Temperature of Coldest Month**
* BIO12 = Annual Precipitation
* BIO16 = Precipitation of Wettest Quarter**
* BIO17 = Precipitation of Driest Quarter**

```{r}
wrld <- geodata::world(path=".")
plot(envs, 1, xlim = c(-130,-80), ylim = c(20, 50)) 
lines(wrld)
points(aacuta[, c("lon", "lat")], col = 'blue')
```

# Extract values from rasters

```{r}
# extract environmental values at aacuta grid cells
aacuta <- na.omit(aacuta[, c('lon', 'lat')])
locs.vals <- extract(envs, aacuta[, c('lon', 'lat')])
# extract environmental values at occ grid cells
print(paste("Records remaining:", nrow(aacuta)))

bgExt <- ext(apply(aacuta[, c('lon', 'lat')], 2, range)) + 0.5
lines(bgExt, lwd=2, col="red")

# crop the environmental rasters by the background extent shape
envsBgCrop <- crop(envs, bgExt)
# sample random background points
bg.xy <- predicts::backgroundSample(envsBgCrop, 5000)
# convert matrix output to data frame
print(paste("Number of background points:", nrow(bg.xy)))

dat <- rbind(locs.vals, bg.vals)
p <- c(rep(1, nrow(locs.vals)), rep(0, nrow(bg.vals)))
MaxEnt(dat, p)
```


# Run Maxent Model & see the maxent results in a browser
```{r}
me <- MaxEnt(envsBgCrop, aacuta, args=c("-J", "-P"))
me
#partialResponse(me)
plot(me)

```

```{r}
r <- predict(me, envsBgCrop)
plot(r)
points(occs.xy, col=gray.colors(1, alpha=.2), pch = 16, cex = 0.5)
mtext(substitute(paste('Suitability for ', italic('Anas acuta'))))
```


# Evaluate SDM

```{r}
occs.xy <- aacuta[c('lon', 'lat')]
nfolds <- 5
set.seed(543)
fold <- folds(occs.xy, k=nfolds)
auc <- rep(NA, nfolds)

for (i in 1:nfolds) {
  occtest <- occs.xy[fold == i, ]
  occtrain <- occs.xy[fold != i, ]
  metst <- MaxEnt(envsBgCrop, occtrain, args=c("-J", "-P"))
  e <- pa_evaluate(me, p = occtest, a = bg.xy, x = envsBgCrop)
  auc[i] <- e@stats["auc"]  
}  
mean(auc)

plot(e, 'ROC') +
mtext(substitute(paste('Sensitivity vs. 1-specificity for ', italic('Anas acuta'))))
  
threshold(e)

```
